---
title: "SegmentedRegression"
author: "Nicholas K. Dulvy"
date: "`r Sys.Date()`"
output: html_document
---

## Running

Open the **repository root** (the `.Rproj`) and render this file from there. In **OffspringSizeRmax** the path is **`Rscripts/depth_temp_analysis/04_DepthTempAnalysis.Rmd`**; in **batoid-rmax-offspring-scaling** it is **`depth_temp_analysis/04_DepthTempAnalysis.Rmd`**. Inputs are resolved from **`data/`** or **`data/processed/`**; figures go to **`figures/`** unless a **`depth_temp_analysis/figures/`** directory exists at the repo root (used in the batoid repo).

```{r setup-knitr-root, include=FALSE}
# Repo root for paths (this file lives under Rscripts/depth_temp_analysis/)
library(here)
knitr::opts_knit$set(root.dir = here::here())
```

```{r}
library(segmented)
library(dplyr)
library(tibble)
library(ape)
library(broom)
library(ggplot2)
library(Cairo)
library(here)

```

```{r paths-data}
# Resolve inputs for either repo layout (batoid: data/*.tree, data/*.csv; OffspringSizeRmax: data/processed/…)
tree_candidates <- c(
  here("data/stein-et-al-single.tree"),
  here("data/processed/stein-et-al-single.tree")
)
tree_path <- tree_candidates[file.exists(tree_candidates)][1]
if (length(tree_path) == 0 || is.na(tree_path)) {
  stop("Missing stein-et-al-single.tree (expected under data/ or data/processed/).")
}

csv_candidates <- c(
  here("data/batoid_model_offspring.csv"),
  here("data/processed/batoid_model_offspring.csv"),
  here("data/processed/Batoid_model_offspring.csv")
)
csv_path <- csv_candidates[file.exists(csv_candidates)][1]
if (length(csv_path) == 0 || is.na(csv_path)) {
  stop("Missing batoid offspring model CSV (expected under data/ or data/processed/).")
}

# Figures: batoid uses depth_temp_analysis/figures/ when present; else repo figures/
figure_dir <- if (dir.exists(here("depth_temp_analysis", "figures"))) {
  here("depth_temp_analysis", "figures")
} else {
  here("figures")
}
dir.create(figure_dir, recursive = TRUE, showWarnings = FALSE)
```

```{r load data}
# Phylogeny loaded for parity with full analysis (not used in chunks below yet)
phy <- ape::read.tree(tree_path)
dat <- read.csv(csv_path, stringsAsFactors = FALSE, header = TRUE)
# select data for models
dat1 <- dat[,c(1, 4,5, 8:15, 24:25)]

#for plotting change genus_species to "genus species"
dat1 <- dat1 %>%
  mutate(
    SciName = gsub("_", " ", BinomName),
    SciName = recode(
      SciName,
      "Manta birostris" = "Mobula birostris",
      "Torpedo californica" = "Tetronarce californica"
    )
  ) %>%
  relocate(SciName, .after = 1)

```

```{r fit segmented models per order}
# Prepare the data
dat1_clean <- dat1 %>%
  filter(!is.na(depth_median), !is.na(t_median)) %>%
  group_by(Order) %>%
  filter(n() >= 10) %>%  # Ensure enough data per group
  ungroup()

mrays   <- lm(depth_median ~ t_median, data = dat1_clean %>% filter(Order == "rays"))
mskates <- lm(depth_median ~ t_median, data = dat1_clean %>% filter(Order == "skates"))

segrays   <- segmented(mrays, seg.Z = ~ t_median, npsi = 1)
segskates <- segmented(mskates, seg.Z = ~ t_median, npsi = 1)

AIC(mrays, segrays)
AIC(mskates, segskates)

# Raw data
rays_data   <- dat1_clean %>% filter(Order == "rays")
skates_data <- dat1_clean %>% filter(Order == "skates")

# Augment with predictions
rays_pred_linear    <- augment(mrays)
rays_pred_segmented <- augment(segrays)
skates_pred_linear    <- augment(mskates)
skates_pred_segmented <- augment(segskates)

# Add model labels
rays_pred_linear$model    <- "Linear"
rays_pred_segmented$model <- "Segmented"
skates_pred_linear$model    <- "Linear"
skates_pred_segmented$model <- "Segmented"

# Extract breakpoints
rays_bp <- segrays$psi[1, "Est."]
skates_bp <- segskates$psi[1, "Est."]

# Combine for plotting
rays_plotdata <- bind_rows(rays_pred_linear, rays_pred_segmented)
skates_plotdata <- bind_rows(skates_pred_linear, skates_pred_segmented)

ggplot(rays_data, aes(x = t_median, y = depth_median)) +
  geom_point(alpha = 0.4) +
  geom_line(data = rays_plotdata, aes(x = t_median, y = .fitted, color = model), linewidth = 1.2) +
  geom_vline(xintercept = rays_bp, linetype = "dotted", color = "red") +
  scale_y_reverse() +
  labs(title = "Rays: Depth vs Temperature",
       x = "Median Temperature (°C)",
       y = "Median Depth (m)") +
  theme_minimal()

ggplot(skates_data, aes(x = t_median, y = depth_median)) +
  geom_point(alpha = 0.4) +
  geom_line(data = skates_plotdata, aes(x = t_median, y = .fitted, color = model), linewidth = 1.2) +
  geom_vline(xintercept = skates_bp, linetype = "dotted", color = "red") +
  scale_y_reverse() +
  labs(title = "Skates: Depth vs Temperature",
       x = "Median Temperature (°C)",
       y = "Median Depth (m)") +
  theme_minimal()


```


```{r}
# Add Order info to predictions
rays_pred_segmented$Order <- "rays"
skates_pred_segmented$Order <- "skates"
segmented_preds <- bind_rows(rays_pred_segmented, skates_pred_segmented)

# Add Order to raw data
rays_data$Order <- "rays"
skates_data$Order <- "skates"
raw_data <- bind_rows(rays_data, skates_data)

# Breakpoints
rays_bp <- segrays$psi[1, "Est."]
skates_bp <- segskates$psi[1, "Est."]

# Plot
ggplot(raw_data, aes(x = t_median, y = depth_median, color = Order)) +
  geom_point(alpha = 0.5, size = 3) +
  geom_line(data = segmented_preds,
            aes(x = t_median, y = .fitted, color = Order),
            linewidth = 1.2) +
  geom_vline(xintercept = rays_bp, color = "darkred", linetype = "dotted") +
  geom_vline(xintercept = skates_bp, color = "blue4", linetype = "dotted") +
  scale_y_reverse() +
  labs(
    title = "Segmented Regression of Depth vs Temperature: Rays and Skates",
    x = "Median Temperature (°C)",
    y = "Median Depth (m)"
  ) +
  theme_minimal(base_size = 14) +
  theme(legend.position = "right")
```

```{r final temp-depth plot}

#
## =========================================================
## 0) One knob: how close the leader line should stop before the text
##    (in x-units = °C)
## =========================================================
label_gap <- 0.15

## =========================================================
## 1) Data + Group (Skates top, Rays bottom)
## =========================================================
dat_plot <- dat1 %>%
  mutate(
    Group = case_when(
      grepl("skate", tolower(Order)) ~ "Skates",
      grepl("ray",   tolower(Order)) ~ "Rays",
      TRUE ~ NA_character_
    )
  ) %>%
  filter(!is.na(Group)) %>%
  mutate(Group = factor(Group, levels = c("Skates", "Rays")))

## =========================================================
## 2) Mean temperature per Group (line + bottom annotation)
##    - Rays label left of line; skates label right of line
## =========================================================
mean_temp2 <- dat_plot %>%
  group_by(Group) %>%
  summarise(mean_t = mean(t_median, na.rm = TRUE), .groups = "drop") %>%
  mutate(
    label     = paste0("Mean temperature = ", round(mean_t, 1), "°C"),
    x_pos     = ifelse(Group == "Rays", mean_t - 0.6, mean_t + 0.6),
    hjust_val = ifelse(Group == "Rays", 1, 0),
    y_pos     = 1.2
  )

## =========================================================
## 3) Select species to label
##    - 2 hottest skates
##    - 5 cold+deep rays (rank-sum)
## =========================================================
skates2_hot <- dat1 %>%
  filter(grepl("skate", tolower(Order))) %>%
  mutate(
    Group = "Skates",
    rank_hot = rank(-t_median, ties.method = "first")
  ) %>%
  arrange(rank_hot) %>%
  slice_head(n = 2) %>%
  select(SciName, Group, t_median, depth_median)

rays5 <- dat1 %>%
  filter(grepl("ray", tolower(Order))) %>%
  mutate(
    Group = "Rays",
    rank_temp  = rank(t_median, ties.method = "first"),
    rank_depth = rank(-depth_median, ties.method = "first"),
    rank_sum   = rank_temp + rank_depth
  ) %>%
  arrange(rank_sum) %>%
  slice_head(n = 5) %>%
  select(SciName, Group, t_median, depth_median)

to_label <- bind_rows(skates2_hot, rays5)

## =========================================================
## 4) Deterministic label positions (species-specific offsets)
##    - x_offset is in °C
##    - y_offset is proportional (works well on log10 depth axis):
##        y_lab = depth * (1 + y_offset)
##    - leader line ends just before the text, on the side facing the point
## =========================================================
to_label <- to_label %>%
  mutate(
    # --- horizontal offsets (°C) ---
    x_offset = case_when(
      SciName == "Urolophus expansus"  ~  1.0,
      SciName == "Tetronarce californica" ~  -1.0,
      SciName == "Mobula birostris"    ~  -1.0,
      SciName == "Mobula mobular"      ~  1.0,
      SciName == "Torpedo marmorata"   ~  -1.0,
      SciName == "Raja eglanteria"     ~  -1.0,
      SciName == "Raja asterias"       ~  -1.0,
      TRUE                             ~  0.5
    ),

    # --- vertical offsets (proportional) ---
    y_offset = case_when(
      SciName == "Urolophus expansus"  ~  0.30,
      SciName == "Tetronarce californica" ~ -0.30,
      SciName == "Mobula birostris"    ~  0.30,
      SciName == "Mobula mobular"      ~  0.30,
      SciName == "Torpedo marmorata"   ~  0.30,
      SciName == "Raja eglanteria"     ~  2.00,
      SciName == "Raja asterias"       ~ -0.30,
      TRUE                             ~  0.00
    ),

    # --- label positions ---
    x_lab = t_median + x_offset,
    y_lab = depth_median * (1 + y_offset),

    # --- label alignment: if label is to the right, left-justify; if to the left, right-justify
    hjust_lab = ifelse(x_lab >= t_median, 0, 1),

    # --- leader line ends just BEFORE the text (so it doesn't start at the text anchor)
    label_dir = ifelse(x_lab >= t_median, 1, -1),
    x_seg_end = x_lab - label_dir * label_gap,
    y_seg_end = y_lab
  )

## =========================================================
## 5) Base plot
## =========================================================
pp_stack <- ggplot(dat_plot, aes(x = t_median, y = depth_median, color = rmax)) +
  geom_point(alpha = 0.7, size = 1.4) +
  geom_smooth(method = "loess", se = FALSE, color = "grey25", linetype = "solid") +
  geom_vline(data = mean_temp2, aes(xintercept = mean_t), color = "grey50", linewidth = 0.6) +
  geom_text(
    data = mean_temp2,
    aes(x = x_pos, y = y_pos, label = label, hjust = hjust_val),
    inherit.aes = FALSE,
    size = 2.0,
    color = "grey55"
  ) +
  facet_grid(Group ~ ., scales = "fixed") +
  scale_y_log10(
    limits = c(1, 1500),
    breaks = c(10, 100, 1000),
    minor_breaks = c(2, 5, 20, 50, 200, 500)
  ) +
  coord_cartesian(xlim = c(0, 30)) +
  scale_color_viridis_c(option = "plasma", name = expression(r[max])) +
  labs(x = "Median Temperature (°C)", y = "Median Depth (m)") +
  theme_bw(base_size = 8) +
  theme(
    strip.text = element_text(face = "bold", size = 8),
    axis.title = element_text(size = 8),
    axis.text  = element_text(size = 7),
    legend.title = element_text(size = 8),
    legend.text  = element_text(size = 7),
    panel.grid = element_blank(),
    panel.border = element_rect(color = "black", fill = NA, linewidth = 0.4),
    plot.margin = margin(3, 3, 3, 3),
    legend.position = "right"
  )

## =========================================================
## 6) Add deterministic labels + leader lines (NO ggrepel)
## =========================================================
## =========================================================
## Add deterministic species labels + leader lines
## =========================================================
pp_labeled <- pp_stack +
  geom_segment(
    data = to_label,
    aes(x = t_median, y = depth_median, xend = x_seg_end, yend = y_seg_end),
    inherit.aes = FALSE,
    linewidth = 0.3,
    colour = "grey40"
  ) +
  geom_text(
    data = to_label,
    aes(x = x_lab, y = y_lab, label = SciName, hjust = hjust_lab),
    inherit.aes = FALSE,
    fontface = "italic",
    size = 1.5,
    vjust = 0,
    colour = "black"
  )

## =========================================================
## Add panel letters (a, b) to the top-left of each facet panel
##  - Uses -Inf/Inf so it stays pinned to the corner when resized
##  - Group order follows the facet order in dat_plot (Skates, then Rays)
## =========================================================
panel_labels <- tibble(
  Group = factor(levels(dat_plot$Group), levels = levels(dat_plot$Group))
) %>%
  mutate(
    lab = ifelse(Group == "Rays", "a", "b"),
    x   = -Inf,
    y   =  Inf
  )

pp_labeled <- pp_labeled +
  geom_text(
    data = panel_labels,
    aes(x = x, y = y, label = lab),
    inherit.aes = FALSE,
    hjust = -0.5,   # nudge right into panel
    vjust =  1.35,   # nudge down into panel
    fontface = "bold",
    size = 3.5,
    colour = "black"
  )

pp_labeled
## =========================================================
## 6) Save: single-column square PNG + PDF
## =========================================================
ggsave(
  file.path(figure_dir, "Depth_Temp_Skates_Rays_singlecol.png"),
  plot = pp_labeled,
  width = 3.4, height = 3.4, units = "in",
  dpi = 600, bg = "white"
)

ggsave(
  file.path(figure_dir, "Depth_Temp_Skates_Rays_singlecol.pdf"),
  plot = pp_labeled,
  width = 3.4, height = 3.4, units = "in",
  device = grDevices::cairo_pdf
)
```



```{r 2 vs 3 segments}

# base lm
moverall <- lm(depth_median ~ t_median, data = dat1_clean)

# 1-breakpoint model (2 segments)
seg_overall_1bp <- segmented(moverall, seg.Z = ~ t_median, npsi = 1)

# 2-breakpoint model (3 segments)
seg_overall_2bp <- segmented(moverall, seg.Z = ~ t_median, npsi = 2)

AIC(moverall, seg_overall_1bp, seg_overall_2bp)



#make predictions
overall_data <- dat1_clean 

# Augment data
fit_lm     <- augment(moverall, newdata = overall_data)
fit_seg1   <- augment(seg_overall_1bp, newdata = overall_data)
fit_seg2   <- augment(seg_overall_2bp, newdata = overall_data)

# Add model labels
fit_lm$model    <- "Linear"
fit_seg1$model  <- "Segmented: 1 Breakpoint"
fit_seg2$model  <- "Segmented: 2 Breakpoints"

# Combine
overall_fits <- bind_rows(fit_lm, fit_seg1, fit_seg2)

qq <- ggplot(overall_data, aes(x = t_median, y = depth_median)) +
  geom_point(alpha = 0.5, size = 2) +
  geom_line(data = overall_fits, aes(x = t_median, y = .fitted, color = model), linewidth = 1.2) +
  scale_y_reverse() +
  labs(
    title = "Depth vs Temperature with Segmented Regression",
    x = "Median Temperature (°C)",
    y = "Median Depth (m)"
  ) +
  theme_minimal(base_size = 14)

```


```{r}

# plot hist for skate and ray depths and temps

# plot facet wrap for temps at different depths
```



